The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-27 script/en/chapters/numbertheory.tex
CT script, de+en: minor corrections

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorization, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and the commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which the modulo operation is performed does not matter.
\end{itemize}
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The property of closeness is always defined in relation to an operation in a set. See 
\hyperlink{Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 = -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{Appendix_C}{Appendix C of this chapter}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{\label{ftn-mod6}Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
For all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix A of this chapter}). Alternatively, we can also use the primitive roots.}.

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{Appendix_A}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem \ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

@@

Alice and Bob want to negotiate a secret session key $S$ via a channel that may be intercepted. 
\begin{itemize}
\item[{\bf 1.}] They select a prime number $p$ and a random number $g$ and exchange this information openly.
\item[{\bf 2.}] Alice now selects $a$, a random number less than $p$ and keeps it secret.

@@

\vskip + 5pt
We will now use an example with (unrealistically) small numbers to illustrate this.
\vskip +1em

@@

Before applying the RSA procedure to a text, we will first demonstrate it
directly using a single number as message\footnote{Using CrypTool\index{CrypTool} you can solve this with the menu 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $11$ and $39$ and must be relatively prime
                to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1{\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

We have the text ``ATTACK AT DAWN'' and the characters are coded in the following simple manner\footnote{Using
CrypTool\index{CrypTool} you can solve this with the menu 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.
This is also described in the tutorial/scenario in CrypTool's online help [Options, specify alphabet, number system, block length\index{Block length} 2 and decimal representation].}:

@@

{\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{Appendix_A}{appendix A of this chapter}}. \\

{\bf 4. Encryption}:\\
{\tt
\begin{tabular}{rcccccccccccccccccccc}
{\rm Text:} & A & T & T & A & C & K & & A & T &  & D & A & W & N \\
{\rm Number:} & 01 & 20 & 20 & 01 & 03 & 11 & 00 & 01 & 20 & 00 & 04 & 01 & 23 & 14
\end{tabular}
}

This $28$-digit number is divided into $4$-digit parts
(because $2626$ is still smaller than $n=3713$):\\
{\tt 0120 2001 0311 0001 2000 0401 2314}

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code
to do RSA encryption using Sage. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

{\bf 5. Decryption}: \\
Cipher text: {\tt 1404 2932 3536 0001 3284 2280 2235 }\\
This $28$-digit number is divided into $4$-digit parts.\\
All 7 parts are decrypted using:  $M \equiv C^{97}{\rm ~(mod~}3713)$: \\
{\tt 0120 2001 0311 0001 2000 0401 2314} \\
The 2-digit numbers are transformed into capital letters and blanks.

Using the selected values it is easy for a cryptanalyst\index{Cryptanalysis}
to derive the secret values from the public 
parameters $n=3713$ and $e=37$ by revealing that $3713 = 47 * 79$.\\  
If $n$ is a $768$-bit number, there is, according to present knowledge, little chance of this.

@@

The idea for this task\footnote{Using CrypTool\index{CrypTool} you can 
solve this via the menu path
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.} 
is taken from the example in \cite[p. 271]{Eckert2003}\index{Eckert 2003}.

Coded in decimal notation, the text ``RSA works!'' is as follows: \\
{\tt
\begin{tabular}{rcccccccccccccccccccc}
{\rm Text:} & R & S & A &   & w & o & r & k & s & ! \\
{\rm Number:} & 82 & 83 & 65 & 32 & 119 & 111 & 114 & 107 & 115 & 33 
\end{tabular} } % \tt

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix E of this chapter} for the source
code to factorize the number $J(n)$ using Sage.
Using CrypTool\index{CrypTool} you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorization of a Number}.}. \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorized but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

The letters are not combined\footnote{For secure procedures we need large numbers that assume -- as far as possible -- all 
values up to $n-1$. If the possible value set for the numbers in the message is too small, even large prime numbers 
cannot make the procedure secure.  An ASCII character is represented by $8$ bits. If we want larger values we must 
combine several numbers. Two characters need $16$ bits, whereby the maximum value that can be represented is $65536$. 
The modulus $n$ must then be greater than $2^{16} = 65536$. This is applied in variant 2.
When the numbers are combined, the leading zeros are kept in binary notation (just as if we were to write all numbers with $3$ 
digits in decimal notation above and were then to obtain the sequence 
{\tt 082 083,  065 032, 119 111,  114 107,  115 033}).}!

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

{\bf 5. Decryption}:\\
Cipher text:

@@

Each character is decrypted using: $M \equiv C^{231,953}{\rm ~mod~}256,027$: \\
{\tt 82 83 65 32 119 111 114 107 115 33}

@@

{\bf 4. Encryption}:

Blocks are formed\footnote{\vskip +3 pt \tt \begin{tabular}{ll@{ }l@{ }l}
single character& binary representation  && decimal representation\\
01010010, 82 & 01010010 01010011 & = &21075 \\
01010011, 83 & \\
01000001, 65 & 01000001 00100000 & = &16672  \\
00100000, 32  \\
01110111, 119 & 01110111 01101111 & = &30575 \\
01101111, 111 \\ 
01110010, 114 & 01110010 01101011 & = &29291 \\
01101011, 107 \\
01110011, 115 & 01110011 00100001 & = &29473 \\
00100001, 33: 
\end{tabular}} (each ASCII character is encoded into a 8 digit binary number below):\\
{\tt 21075 16672 30575 29291 29473}\footnote{Using CrypTool\index{CrypTool} 
you can solve this with the menu
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration} 
with the following options: all 256 ASCII characters, b-adic, block length\index{Block length} 2 and decimal representation.}

\label{SrcArith4d}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 158721 137346 37358 240130 112898}

{\bf 5. Decryption:} \\
Cipher text:\\
{\tt 158721 137346 37358 240130 112898}

Each block is decrypted using: $M \equiv C^{231,953}{\rm ~(mod~}256,027)$: \\
{\tt 21075 16672 30575 29291 29473}

@@

{\bf 4'. Encryption:} \\
Blocks are formed: (each ASCII character is encoded into a 3 digit decimal number below):\\
 {\tt 82083 65032 119111 114107 115033}\footnote{The RSA encryption works correctly with the
modulus $n=256.027$ because each ASCII block of two characters will be encoded into a number that is smaller or equal than
the number $255,255$.  }

\label{SrcArith4e}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 198967 051405 254571 115318 014251}

{\bf 5'. Decryption:} \\
Cipher text:\\
 {\tt 198967 051405 254571 115318 014251}

Each block is decrypted using: $M \equiv C^{2473}{\rm ~(mod~}67,519)$: \\
{\tt 82083 65032 119111 114107 115033}

@@

$n = 18,923$ and $e = 1261$ (for Table 4.1) and \\
$n = 31,313$ and $e = 4913$ (for Table 4.2).

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_B}{}
\section*{Appendix B: Forming closed sets}
\addcontentsline{toc}{section}{Appendix B: Forming closed sets}

>>>>

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{NumberTheory_Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorization, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and the commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which the modulo operation is performed does not matter.
\end{itemize}
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The property of closeness is always defined in relation to an operation in a set. See 
\hyperlink{NumberTheory_Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 = -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{NumberTheory_Appendix_C}{Appendix C of this chapter}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{\label{ftn-mod6}Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
For all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{NumberTheory_Appendix_A}{appendix A of this chapter}). Alternatively, we can also use the primitive roots.}.

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{NumberTheory_Appendix_A}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem \ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

@@

\noindent Alice and Bob want to negotiate a secret session key $S$ via a channel that may be intercepted. 
\begin{itemize}
\item[{\bf 1.}] They select a prime number $p$ and a random number $g$ and exchange this information openly.
\item[{\bf 2.}] Alice now selects $a$, a random number less than $p$ and keeps it secret.

@@

\vskip + 5pt
\noindent We will now use an example with (unrealistically) small numbers to illustrate this.
\vskip +1em

@@

Before applying the RSA procedure to a text, we will first demonstrate it
directly using a single number as message\footnote{Using CrypTool\index{CrypTool} you can solve this with the menu 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $11$ and $39$ and must be relatively prime
                to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1{\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

We have the text ``ATTACK AT DAWN'' and the characters are coded in the following simple manner\footnote{Using
CrypTool\index{CrypTool} you can solve this with the menu 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.
This is also described in the tutorial/scenario in CrypTool's online help [Options, specify alphabet, number system, block length\index{Block length} 2 and decimal representation].}:

@@

\noindent {\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3713; ~J(n) = (p-1)*(q-1)=3588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3587$ and must be relatively prime
                to $3588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3589 \equiv 1{\rm ~(mod~}3588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{NumberTheory_Appendix_A}{appendix A of this chapter}}. \\

\noindent {\bf 4. Encryption}:\\
{\tt
\begin{tabular}{rcccccccccccccccccccc}
{\rm Text:} & A & T & T & A & C & K & & A & T &  & D & A & W & N \\
{\rm Number:} & 01 & 20 & 20 & 01 & 03 & 11 & 00 & 01 & 20 & 00 & 04 & 01 & 23 & 14
\end{tabular}
}

\noindent This $28$-digit number is divided into $4$-digit parts
(because $2626$ is still smaller than $n=3713$):\\
{\tt 0120 2001 0311 0001 2000 0401 2314}

\label{SrcArith4a}
\noindent All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code
to do RSA encryption using Sage. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

\noindent {\bf 5. Decryption}: \\
Cipher text: {\tt 1404 2932 3536 0001 3284 2280 2235 }

\noindent This $28$-digit number is divided into $4$-digit parts.

\noindent All 7 parts are decrypted using:  $M \equiv C^{97}{\rm ~(mod~}3713)$: \\
{\tt 0120 2001 0311 0001 2000 0401 2314}

\noindent The 2-digit numbers are transformed into capital letters and blanks.

\noindent Using the selected values it is easy for a cryptanalyst\index{Cryptanalysis}
to derive the secret values from the public 
parameters $n=3713$ and $e=37$ by revealing that $3713 = 47 * 79$.

\noindent If $n$ is a $768$-bit number, there is, according to present knowledge, little chance of this.

@@

\noindent The idea for this task\footnote{Using CrypTool\index{CrypTool} you can 
solve this via the menu path
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.} 
is taken from the example in \cite[p. 271]{Eckert2003}\index{Eckert 2003}.

\noindent Coded in decimal notation, the text ``RSA works!'' is as follows: \\
{\tt
\begin{tabular}{rcccccccccccccccccccc}
{\rm Text:} & R & S & A &   & w & o & r & k & s & ! \\
{\rm Number:} & 82 & 83 & 65 & 32 & 119 & 111 & 114 & 107 & 115 & 33 
\end{tabular} } % \tt

\noindent We will work through the example in 2 variants. The steps 1 to 3 are common for both.
\par
\noindent {\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix E of this chapter} for the source
code to factorize the number $J(n)$ using Sage.
Using CrypTool\index{CrypTool} you can solve this with the 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorization of a Number}.}. \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorized but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\noindent The letters are not combined\footnote{For secure procedures we need large numbers that assume -- as far as possible -- all 
values up to $n-1$. If the possible value set for the numbers in the message is too small, even large prime numbers 
cannot make the procedure secure.  An ASCII character is represented by $8$ bits. If we want larger values we must 
combine several numbers. Two characters need $16$ bits, whereby the maximum value that can be represented is $65536$. 
The modulus $n$ must then be greater than $2^{16} = 65536$. This is applied in variant 2.
When the numbers are combined, the leading zeros are kept in binary notation (just as if we were to write all numbers with $3$ 
digits in decimal notation above and were then to obtain the sequence 
{\tt 082 083,  065 032, 119 111,  114 107,  115 033}).}!

\label{SrcArith4c}
\noindent Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

\noindent {\bf 5. Decryption}:\\
Cipher text:

@@

\noindent Each character is decrypted using: $M \equiv C^{231,953}{\rm ~mod~}256,027$: \\
{\tt 82 83 65 32 119 111 114 107 115 33}

@@

\noindent {\bf 4. Encryption:}\\
Blocks are formed\footnote{\vskip +3 pt \tt \begin{tabular}{ll@{ }l@{ }l}
single character& binary representation  && decimal representation\\
01010010, 82 & 01010010 01010011 & = &21075 \\
01010011, 83 & \\
01000001, 65 & 01000001 00100000 & = &16672  \\
00100000, 32  \\
01110111, 119 & 01110111 01101111 & = &30575 \\
01101111, 111 \\ 
01110010, 114 & 01110010 01101011 & = &29291 \\
01101011, 107 \\
01110011, 115 & 01110011 00100001 & = &29473 \\
00100001, 33: 
\end{tabular}} (each ASCII character is encoded into a 8 digit binary number below):\\
{\tt 21075 16672 30575 29291 29473}\footnote{Using CrypTool\index{CrypTool} 
you can solve this with the menu
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration} 
with the following options: all 256 ASCII characters, b-adic, block length\index{Block length} 2 and decimal representation.}

\label{SrcArith4d}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 158721 137346 37358 240130 112898}

\noindent {\bf 5. Decryption:} \\
Cipher text:\\
{\tt 158721 137346 37358 240130 112898}

\noindent Each block is decrypted using: $M \equiv C^{231,953}{\rm ~(mod~}256,027)$: \\
{\tt 21075 16672 30575 29291 29473}

@@

\noindent {\bf 4'. Encryption:} \\
Blocks are formed: (each ASCII character is encoded into a 3 digit decimal number below):\\
 {\tt 82083 65032 119111 114107 115033}\footnote{The RSA encryption works correctly with the
modulus $n=256.027$ because each ASCII block of two characters will be encoded into a number that is smaller or equal than
the number $255,255$.  }

\label{SrcArith4e}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 198967 051405 254571 115318 014251}

\noindent {\bf 5'. Decryption:} \\
Cipher text:\\
 {\tt 198967 051405 254571 115318 014251}

\noindent Each block is decrypted using: $M \equiv C^{2473}{\rm ~(mod~}67,519)$: \\
{\tt 82083 65032 119111 114107 115033}

@@

\noindent $n = 18,923$ and $e = 1261$ (for Table 4.1) and \\
\noindent $n = 31,313$ and $e = 4913$ (for Table 4.2).

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{NumberTheory_Appendix_B}{}
\section*{Appendix B: Forming closed sets}
\addcontentsline{toc}{section}{Appendix B: Forming closed sets}
